---
title: "01_credit_risk_eda"
author: "HJ Pink"
date: "2023-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Credit Risk - merge and combine datasets

<https://www.kaggle.com/competitions/home-credit-default-risk/data>

## Business Problem

Home Credit Group addresses the challenge of limited access to loans for individuals with insufficient credit histories. Many people face difficulties in obtaining loans, which makes them vulnerable to unreliable lenders. Home Credit strives to provide a positive and secure borrowing experience for this underserved population. They leverage alternative data sources, such as telco and transactional information, to assess clients' repayment abilities.

By utilizing statistical and machine learning techniques, Home Credit aims to unlock the full potential of their data and make accurate predictions. Their goal is to ensure that deserving individuals are not denied loans and to provide them with suitable loan terms and repayment schedules. Through this approach, Home Credit contributes to greater financial inclusion and empowers clients to achieve financial success.

### Load libraries

```{r message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(magrittr)
library(DT)

##import helper functions
source('helper_functions.R')
```

## Import Data

This dataset contains a relational schema of several tables. `application_{train|test}.csv` contains main information regarding the credit application including a primary key (`sk_id_curr`). applications_train contains 1 row per `sk_id_curr`. However other available datasets require aggreagtion before merging, including; bank balances, previous credit applications_train, etc contain information regarding previous loans (successful and unsuccessful), credit card spending, repayment history, etc

### Application data

Import `application_train.csv`, this is the main table which contains data regarding the credit application including the **target variable - whether or not the loan defaulted**.

Create additional features such as income ratios and discretise continuous features into binary (does applicant have children).

```{r}
read_applications <- function(path){
  df <- clean_read_csv(path) %>%
  mutate(
    #convert days_birth to age
    age = abs(days_birth)/365.25,
    age_group = cut(age, 
                    breaks = c(0, 30, 45, 60, Inf),
                    labels = c("under 30", "30-45", "45-60", "60+"),
                    include.lowest = TRUE) %>% factor(),
    yrs_employed = abs(days_employed)/365.25,
    yrs_employed = ifelse(yrs_employed > age,
                           age*median(yrs_employed/age), 
                           yrs_employed),
    # Calculate the credit-to-income ratio
    credit_income_ratio = amt_credit / amt_income_total,
    # Calculate the annuity-to-income ratio
    annuity_income_ratio = amt_annuity / amt_income_total,
    # Calculate income per family member
    income_per_family_member = amt_income_total / cnt_fam_members, 
    # Calculate the percentage of adult life employed
    perc_adult_life_employed = yrs_employed / (age-16),
    # Clean and factorize the organization type column
    organization_type = organization_type %>% 
      gsub('Type \\d{1,2}|: type \\d{1,2}', '', .) %>% 
      trimws() %>% factor(),
    # Create a binary variable for childless applicants
    is_childless = ifelse(cnt_children == 0, 1, 0) %>% factor()
  ) %>% 
  select(-days_employed, -days_birth)
  
  return(df)
    
}
applications_train <- read_applications('data/application_train.csv') 
applications_test <- read_applications('data/application_test.csv')

dt_datatable_wrapper(applications_train)
```

### Bureau data

Import data credit bureau regarding the previous loans including amount credit available to each application, amount of debt to be re-payed, whether payments were made late.

```{r}

#C means closed, X means status unknown, 0 means no DPD, 1 means maximal did during month between 1-30, 2 means DPD 31-60,… 5 means DPD 120+ or sold or written off ]
bureau_balance <- clean_read_csv('data/bureau_balance.csv')
balance_status_to_numeric <- c(
  'C' = 0,'X' = 0, '0' = 0, '1' = 1, '2' = 2, '3' = 3, '4' = 4, '5' = 5
)
bureau_balance$status_numeric <- plyr::mapvalues(bureau_balance$status,
                                           from = names(balance_status_to_numeric),
                                           to = balance_status_to_numeric) %>%
  as.numeric()

bureau_balance$months_balance <- abs(bureau_balance$months_balance)
bureau_balance$weighted_status <- bureau_balance$status_numeric/(bureau_balance$months_balance+1)




previous_credits <- clean_read_csv('data/bureau.csv') %>%
  # Replace NA values with zeros to avoid issues with ratios
  replace_na(list(
    amt_credit_sum = 0, 
    amt_credit_sum_debt = 0,
    amt_credit_sum_overdue = 0,
    credit_day_overdue = 0, 
    amt_annuity = 0
  )) %>%
  mutate(
    # Calculate days since credit update
    days_since_credit_update = -days_credit_update,
    # Calculate the debt-to-credit ratio
    debt_to_credit_ratio = ifelse(amt_credit_sum == 0, 0, amt_credit_sum_debt / amt_credit_sum),
    # Calculate the overdue-to-credit ratio
    overdue_to_credit_ratio = ifelse(amt_credit_sum == 0, 0, amt_credit_sum_overdue / amt_credit_sum)
  )

dt_datatable_wrapper(previous_credits)

```

### Credit card data

The credit_card_balance.csv dataset contains information about monthly credit card balances for previous credits in Home Credit. It includes data such as the ID of previous credits, loan IDs, months of balance relative to the application date, balance amounts, credit card limits, amounts drawn, payment information, receivables, installment counts, contract statuses, and days past due.

The dataset provides insights into the credit card usage and payment behaviour of borrowers. It can be used to analyse trends in credit card balances, identify default patterns, and assess the creditworthiness of borrowers.

```{r}
credit_card_balances <- clean_read_csv('data/credit_card_balance.csv') %>%
  # Replace NAs with zero
  replace_na(list(
    amt_drawings_atm_current = 0, 
    amt_drawings_current = 0,
    amt_drawings_other_current = 0, 
    amt_drawings_pos_current = 0
  )) %>% 
  # Calculate the total drawings
  mutate(
    total_drawings = amt_drawings_atm_current + 
                     amt_drawings_current + 
                     amt_drawings_other_current + 
                     amt_drawings_pos_current,
    # Create a binary variable for delayed payments
    dpd_delayed = ifelse(sk_dpd > 0, 1, 0), 
    months_balance = abs(months_balance)
  )


dt_datatable_wrapper(credit_card_balances)
```

### Installment payment data

The `installments_payments.csv` dataset contains data on instalment payments for previous credits in Home Credit. It includes information such as previous credit IDs, loan IDs, instalment versions, numbers, due and payment days, prescribed and actual payment amounts. The dataset helps track payment behaviors, assess repayment patterns, and evaluate creditworthiness. It provides insights into borrowers' payment discipline and fulfilment of instalment obligations. This information is valuable for analyzing payment trends, predicting default risks, and optimizing loan management.

```{r}
installments_payments <- clean_read_csv('data/installments_payments.csv') %>% 
  ##replace NAs with zeros
  replace_na(list(days_entry_payment =0, amt_payment =0)) %>% 
  ##calculate new features based on existing data 
  mutate(
    amt_instalment = ifelse(amt_instalment==0, 0.1, amt_instalment),
    #when was the instalment paid relative to due date
    days_before_deadline =days_instalment-days_entry_payment,
    #binary feature for if instalment payment was late
    is_late = ifelse(days_before_deadline <0, 1, 0), 
    #how much of instalment value was paid
    perc_install_paid = amt_payment/amt_instalment, 
    #binary for if instalment wasn't fully paid
    is_underpaid = ifelse(perc_install_paid <1, 1,0 )
  )
  
dt_datatable_wrapper(installments_payments)

```

### Previous credit application data

The `previous_application.csv` dataset contains information about previous loan applications_train in Home Credit. It includes data such as IDs of previous credits and loans, contract details, payment amounts, application timing, flags indicating the last application per contract or day, interest rates, loan purposes, decision days, payment methods, rejection reasons, client types, goods categories, portfolio and product types, channel types, seller details, payment terms, and insurance requests. This dataset provides valuable insights into loan application patterns, client preferences, and application outcomes. It can be utilized for risk assessment, customer segmentation, and improving loan application processes.

```{r}
previous_applications <- clean_read_csv('data/previous_application.csv') %>% 
  mutate(is_downpayment = ifelse(!is.na(amt_down_payment) & amt_down_payment>0, 
                                 1,0))

##filter based on success of application

accepted_applications <- filter(previous_applications, 
                                name_contract_status %in% c('Approved','Unused offer'))

rejected_applications <- filter(previous_applications, 
                                name_contract_status %in% c('Refused','Canceled'))

dt_datatable_wrapper(rejected_applications)

```

## Data Aggregation

Calculate aggregated statistics on `previous_credits`, `installments_payments`, `credit_card_balances` and `previous_applications` at the loan applicant level (`sk_id_curr`).

### Previous Credit Aggregation

```{r}
bureau_balance_agg <- bureau_balance %>% 
  dplyr::group_by(sk_id_bureau) %>% 
  summarise(
    median_balance_status = median(status_numeric, na.rm=TRUE),
    max_balance_status = max(status_numeric, na.rm=TRUE),
    median_weighted_status = median(weighted_status, na.rm=TRUE),
    max_weighted_status = max(weighted_status, na.rm=TRUE)
  )

rm(bureau_balance)
```


```{r}

bureau_merged <- merge(
  previous_credits ,bureau_balance_agg, 
  by='sk_id_bureau',all.x=TRUE 
)

previous_credits_agg <- bureau_merged %>%
  mutate(days_since_credit_update = abs(days_since_credit_update),
         days_credit = abs(days_credit),
         days_credit_enddate = abs(days_credit_enddate)) %>% 
  group_by(sk_id_curr) %>%
  summarise(
    #do they have car loan
    count_loan_types = n_distinct(credit_type),
    # Calculate the total sum of credit amounts
    total_credit_sum = sum(amt_credit_sum, na.rm = TRUE),
    # Calculate the total sum of credit debts
    total_credit_debt = sum(amt_credit_sum_debt, na.rm = TRUE),
    # Calculate the total sum of coverdue credit
    total_credit_overdue = sum(amt_credit_sum_overdue, na.rm = TRUE),
    total_credit_max_overdue = sum(amt_credit_max_overdue, na.rm=TRUE),
    # Calculate the total sum of days overdue
    total_days_overdue = sum(credit_day_overdue, na.rm = TRUE),
    # Calculate the total sum of annuities
    total_annuity = sum(amt_annuity, na.rm = TRUE),
    # Count the number of active credits
    count_active_credits = sum(credit_active == 'Active'),
    # Count the number of closed credits
    count_closed_credits = sum(credit_active == 'Closed'),
    # Find the minimum debt-to-credit (credit utilisation) ratio
    min_debt_to_credit_ratio = min(debt_to_credit_ratio, na.rm = TRUE),
    # Find the maximum debt-to-credit (credit utilisation) ratio
    max_debt_to_credit_ratio = max(debt_to_credit_ratio, na.rm = TRUE),
    # Find the maximum overdue-to-credit ratio
    max_overdue_to_credit_ratio = max(overdue_to_credit_ratio, na.rm = TRUE),
    # Calculate the average days since credit update
    avg_days_since_credit_update = median(days_since_credit_update, na.rm = TRUE),
    ## fewest days since update 
    min_days_since_credit_update = min(days_since_credit_update, na.rm = TRUE),
    credit_duration = sum(days_credit-days_credit_enddate,na.rm=TRUE) ) %>% 
  mutate(
    # Calculate the overall credit-to-annuity ratio
    credit_to_annuity_ratio = ifelse(total_annuity == 0, 0, 
                                     total_credit_sum / total_annuity),
    # Calculate overall the debt-to-credit ratio
    debt_to_credit_ratio = ifelse(total_credit_sum == 0, 0, 
                                  total_credit_debt / total_credit_sum),
    # Calculate the overall overdue-to-credit ratio
    overdue_to_credit_ratio = ifelse(total_credit_sum == 0, 0, 
                                     total_credit_overdue / total_credit_sum)
  )

dt_datatable_wrapper(previous_credits_agg)

```

### Credit card balance aggregation

```{r}
# Aggregate credit card balances data
credit_data_agg <- credit_card_balances %>%
  mutate(weighting_month = abs(months_balance)+1) %>% 
  group_by(sk_id_curr) %>%
  summarise(
    # Calculate the number of months with balances
    n_months_balances = n(),
    # Calculate the mean balance
    avg_balance = median(amt_balance, na.rm = TRUE),
    mean_weighted_balanced = median(amt_balance/weighting_month),
    # Calculate max balance 
    max_balance = max(amt_balance, na.rm =TRUE), 
    max_weighted_balance = max(amt_balance/weighting_month, na.rm = TRUE),
    # Calculate the standard deviation of balance
    sd_balance = sd(amt_balance, na.rm = TRUE),
    # Calculate the maximum credit limit
    max_credit_limit = max(amt_credit_limit_actual, na.rm = TRUE),
    # Calculate the total sum of drawings
    total_drawings_sum = sum(total_drawings, na.rm = TRUE),
    total_weighted_drawings_sum = sum(total_drawings/weighting_month, na.rm=TRUE),
    # Calculate the maximum monthly drawings
    max_mth_drawings = max(total_drawings, na.rm = TRUE),
    max_weighted_mth_drawings = max(total_drawings/weighting_month, na.rm=TRUE),
    # Calculate the total sum of point-of-sale drawings
    total_point_of_sale_drawings = sum(amt_drawings_pos_current, na.rm = TRUE),
    total_weighted_point_of_sale_drawings = sum(amt_drawings_pos_current/weighting_month, 
                                                na.rm = TRUE),
    # Calculate total inst_min_regularity
    toatl_inst_min_regularity = sum(amt_inst_min_regularity, na.rm = TRUE),
    # Calculate average point of sale drawings
    average_point_of_sale_drawings = median(amt_drawings_pos_current, na.rm = TRUE),
    # Calculate the total sum of payments
    total_payments_sum = sum(amt_payment_total_current, na.rm = TRUE),
    total_weighted_payments_sum = sum(amt_payment_total_current/weighting_month, na.rm = TRUE),
    # Calculate the total sum of payments
    total_payments_max = max(amt_payment_total_current, na.rm = TRUE),
    total_weighted_payments_max = sum(amt_payment_total_current/weighting_month, na.rm = TRUE),
    # Calculate total receivable amount
    total_receivable_sum = sum(amt_total_receivable, na.rm = TRUE),
    total_weighted_receivable_sum = sum(amt_total_receivable/weighting_month, na.rm = TRUE),
    # Calculate median receivable amount
    avg_receivable_sum = median(amt_total_receivable, na.rm = TRUE),
    avg_weighted_receivable_sum = median(amt_total_receivable/weighting_month, na.rm = TRUE),
    # Calculate the maximum DPD (Days Past Due)
    total_dpd = max(sk_dpd, na.rm = TRUE),
    # Calculate the count of delayed payments
    count_dpd_delayed = sum(dpd_delayed, na.rm = TRUE)
  )

dt_datatable_wrapper(credit_data_agg)

```

### Credit card monthly balance change stats

```{r}
# Filter for customers with multiple balances
multiple_balances <- filter(credit_data_agg, n_months_balances > 2)$sk_id_curr 

# Calculate monthly balance change
mth_balance_change <- credit_card_balances %>%
  filter(sk_id_curr %in% multiple_balances) %>% 
  arrange(sk_id_curr, months_balance) %>%
  group_by(sk_id_curr) %>%
  mutate(balance_change = amt_balance - lag(amt_balance))

# Calculate aggregate statistics of balance change
balance_change_agg <- mth_balance_change %>%
  ungroup() %>% 
  group_by(sk_id_curr) %>%
  summarise(
    min_balance_change = min(balance_change, na.rm = TRUE),
    max_balance_change = max(balance_change, na.rm = TRUE), 
    median_balance_change = median(balance_change, na.rm = TRUE)
  )

# Clean up large variable to save memory
rm(credit_card_balances)

# Merge credit data aggregation with balance change aggregation
credit_agg <- merge(credit_data_agg, balance_change_agg, 
                    by = 'sk_id_curr', all.x = TRUE)

dt_datatable_wrapper(balance_change_agg)

```

### Credit payment instalment aggregations

```{r}
# Aggregate installments payments data
installments_agg <- installments_payments %>% 
  group_by(sk_id_curr) %>% 
  summarise(
    # Calculate the number of installments
    n_installments = n(),
    # Calculate the number of late payments
    n_late_payments = sum(is_late, na.rm = TRUE),
    # Calculate the number of underpaid installments
    n_underpaid = sum(is_underpaid, na.rm = TRUE),
    # Calculate the total amount of installment
    total_amt_instalment = sum(amt_instalment, na.rm = TRUE),
    # Calculate the total amount of payment
    total_amt_payment = sum(amt_payment, na.rm = TRUE),
    # Calculate the median days before the deadline
    median_days_before_deadline = median(days_before_deadline, na.rm = TRUE),
    # Calculate the minimum days before the deadline
    min_days_before_deadline = min(days_before_deadline, na.rm = TRUE),
    # Calculate the minimum percentage of installment paid
    min_perc_installment_paid = min(perc_install_paid, na.rm = TRUE)
  ) %>% 
  mutate(
    # Calculate the percentage of late payments
    perc_late = n_late_payments / n_installments,
    # Calculate the percentage of underpaid installments
    perc_underpaid = n_underpaid / n_installments, 
    # Calculate the overall percentage paid
    overall_perc_paid = total_amt_payment / total_amt_instalment
  )

head(installments_agg, 10) %>% datatable()


```

### Previous credit applications_train data aggregation

```{r}
# Count the total number of previous applications_train for each SK_ID_CURR
total_applications <- count(previous_applications, sk_id_curr, name = 'total_num_previous_apps')

# Aggregate accepted applications_train data
accepted_agg <- accepted_applications %>% 
  group_by(sk_id_curr) %>% 
  summarise(
    # Count the total number of accepted applications_train
    total_accepted_apps = n(),  
    # Calculate the sum of accepted credit amounts
    total_accepted_credit = sum(amt_credit, na.rm = TRUE),  
    # Calculate the mean downpayment rate for accepted applications_train
    accepted_median_downpayment_rate = median(rate_down_payment, na.rm = TRUE),  
    # Calculate the proportion of accepted applications_train with downpayment
    accepted_prop_w_downpayment = sum(is_downpayment) / n(),  
    # Find the most recent acceptance date
    most_recent_accepted = max(days_decision, na.rm = FALSE),  
    accepted_max_days_last_due = max(days_last_due, na.rm = FALSE),  
    # Find the maximum days to the last due date for accepted applications_train
    accepted_max_days_first_due = max(days_first_due, na.rm = FALSE)  
  )

# Aggregate rejected applications_train data
rejected_agg <- rejected_applications %>% 
  group_by(sk_id_curr) %>% 
  summarise(
     # Count the total number of rejected applications_train
    total_rejected_apps = n(), 
    # Calculate the sum of rejected credit amounts
    total_rejected_credit = sum(amt_credit, na.rm = TRUE),  
    # Calculate the mean downpayment rate for rejected applications_train
    rejected_median_downpayment_rate = median(rate_down_payment, na.rm = TRUE),  
    # Calculate the proportion of rejected applications_train with downpayment
    rejected_prop_w_downpayment = sum(is_downpayment) / n(),  
    # Find the most recent rejection date
    most_recent_rejected = max(days_decision, na.rm = FALSE)  
  )

# Merge the aggregated data
previous_applications_agg <- merge(total_applications, accepted_agg, 
                                   by = 'sk_id_curr', all.x = TRUE) %>% 
  merge(rejected_agg, by = 'sk_id_curr', all.x = TRUE) %>% 
  mutate(
    # Replace NA, NaN, and infinite values with 0
    across(everything(), ~replace(., is.na(.) | is.nan(.) | is.infinite(.), 0)),
    # Calculate the acceptance rate of previous applications_train
    accepted_rate = total_accepted_apps / total_num_previous_apps,
    # Calculate the proportion of credit rejected
    prop_credit_rejected = total_rejected_credit / 
      (total_rejected_credit + total_accepted_credit)
  )
  
# Print the head of the aggregated statistics
head(previous_applications_agg,10) %>% datatable()


```

## Merge datasets

```{r}
# Merge dataframes

# Merge the 'applications_train' and 'previous_credits_agg' dataframes by 'sk_id_curr', keeping all rows from 'applications_train'

merge_and_clean <- function(application_df){
  df <- merge(application_df, 
              previous_credits_agg,
              by = 'sk_id_curr', all.x = TRUE) %>% 
  merge(credit_agg, 
        by = 'sk_id_curr', all.x = TRUE) %>% 
  merge(installments_agg, 
        by = 'sk_id_curr', all.x = TRUE) %>% 
  merge(previous_applications_agg, 
        by = 'sk_id_curr', all.x = TRUE) 
  
  nrow1 <- nrow(df)
  df <- distinct(df, sk_id_curr,.keep_all = TRUE)
  nrow2 <- nrow(df)
  print(paste('Removed', nrow1-nrow2, 'duplicate rows'))

  # Replace NAs in aggregated columns with zeros. 
  #In these   datasets, NA means no data
  cols2replace <- c(colnames(credit_agg), 
                    colnames(installments_agg),
                    colnames(previous_credits_agg), 
                    colnames(previous_applications_agg)) %>%
    unique()

  df[, cols2replace] <- lapply(df[, cols2replace], 
                               function(x) ifelse(is.na(x), 0, x))
  
  return(df)
  
}

df_train <- merge_and_clean(applications_train)
df_test <- merge_and_clean(applications_test)





```


## Write dataset to local file

```{r}
# Write the merged dataframe to a compressed .rds file
write_rds(df_train, 'data/all_data_merged.rds.gz', compress = 'gz')
write.csv(df_train, 'data/all_data_merged.csv', row.names = FALSE)
write.csv(df_test, 'data/all_TEST_data_merged.csv', row.names = FALSE)
```
